---
title: "Capitulo 1"
output: html_notebook
---

# Chapter 1

First step, install the packages we're need it.

```{r}
# Excecute if you need it
# install.packages("diagram")
# install.packages("expm")
# install.packages("markovchain")
```

First, an n x m transition probability matrix should be specified:

tm.name <- matrix(c(p11.value,  p12.value, . . . , pnn.value), nrow=n.value, ncol=n.value, byrow=TRUE)

The plot the diagram with the arrows pointing in the correct direction, the transition matrix must be transposed. This can be done by typing
tr.tm.name <- t(tm.name)

The diagram is drawn using plotmat function. The graph depicts state names
inside boxes and directed lines connecting the boxes labeled by the corresponding transition probabilities. The syntax is:


library(diagram)
plotmat(tr.tm.name, <arguments>)

To compute an n-step transition probability matrix for a specific value of n, use the following code:

library(expm)
nstepmatrix.name <- tm.name %^% n.value

Given the initial distribution, the lines below calclate the uncoditional distribution after n step.

init.p.name <- c(p1.init.vale, p2.init.value,...)
uncond.dist.name <- init.p.name %*% nstepmatrix.name

To determine recurrent and transient classes, abosrbing states, and the limiting distribution, a discrete-time Markov cahin must be created as an object. It can be done as follows:


library(markovchain)

dtmc.name<- new("markovchain", transitionMatrix=tm.name,
states=c("state1.name", "state2.name",...))


Then state characteristic may be obtained as:

recurrentClasses(dtmc.name)
transientClasses(dtmc.name)
absorbingStates(dtmc.name)
steadyStates(dtmc.name)


Note that R doesn't identify reflecting states.

In R, one can compute the period of an irreducible (a single-class) Markov chain only. The syntax is period(dtme.name)

### Example 1.8
The results obtained for the Markov chain from Examples 1.1-1.4 and 1.6 can be produced in R as presented below.

```{r}
#specifying transition probability matrix
tm<- matrix(c(0.7, 0.1, 0.2, 0.0, 0.6, 0.4, 0.5, 0.2, 0.3),
nrow=3, ncol=3, byrow=TRUE)

#transposing transition probability matrix
tm.tr<- t(tm)

#plotting diagram
library(diagram)

plotmat(tm.tr, pos=c(1,2), arr.length=0.3, arr.width=0.1,
box.col="light blue", box.lwd=1, box.prop=0.5, box.size=0.12,
box.type="circle", cex.txt=0.8, lwd=1, self.cex=0.6,
self.shiftx=0.17, self.shifty=-0.01)
```

```{r}
# computing three-step transition probability matrix
library(expm)
print(tm3<-tm%^%3)
```

```{r}
# computing unconditional distribution after three steps
init.p <- c(1/3, 1/3, 1/3)
init.p%*%tm3
```


```{r}
# Creating Markov chain object
library(markovchain)
mc <- new('markovchain', transitionMatrix = tm, states=c('1','2','3'))

# Computing Markov chain characteristics
recurrentClasses(mc)

transientClasses(mc)

absorbingStates(mc)

period(mc)

round(steadyStates(mc), digits=4)
```

## Example 1.9
Consider the Markov chain from Examples 1.5 and 1.7. We run the following R code to produce the diagram, invariant vectors, and to verify the state classification.
```{r}
# specifying transition probability matrix
tm<- matrix(c(0.3,0.7,0,0,0,0,1,0,0,0,0,0,0.5,0,0,0,0,0.5,
0,0,0.6,0,0,0.4,0,0,0,0,0.1,0.9,0,0,0,0,0.7,0.3), nrow=6,
ncol=6, byrow=TRUE)


# transposing transition probability matrix
tm.tr <- t(tm)

# plotting diagram
library(diagram) 
plotmat(tm.tr, arr.length=0.3, arr.width=0.1, box.col="cyan", box.lwd=1, box.prop=0.5, box.size=0.09,
box.type="circle", cex.txt=0.8, lwd=1, self.cex=0.3,
self.arrpos=0.3, self.shiftx=0.09, self.shifty=-0.05)
```

```{r}
#creating Markov chain object
library(markovchain)
mc<- new("markovchain", transitionMatrix=tm, states=c("1",
"2", "3", "4", "5", "6"))
#computing Markov chain characteristics
recurrentClasses(mc)

transientClasses(mc)
```

```{r}
#finding periods of irreducible Markov chains
tm12.ir<- matrix(c(0.3,0.7,1,0), nrow=2, ncol=2, byrow=TRUE)
mc12.ir<- new("markovchain", transitionMatrix=tm12.ir,
states=c("1","2"))
period(mc12.ir)

tm56.ir<- matrix(c(0.1,0.9,0.7,0.3), nrow=2, ncol=2,
byrow=TRUE)
mc56.ir<- new("markovchain", transitionMatrix=tm56.ir,
states=c("5","6"))
period(mc56.ir)

#finding steady-state distribution
round(steadyStates(mc), digits=4)
```